Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2023-02-04 analysis/paper/reviewer_response.md

> *What about human communities that don't live in the region, but that
> send large LNG tankers and other vessel through the region?  The loss
> of sea ice has effects on those communities and activities too.*

There is a seemingly endless list of impacts of loss of sea ice to human
communities (and ecosystems) beyond the Arctic region to the rest of the
world. However, for this sentence, we wanted to keep the reader focused
on the Arctic region.

> *Please use direct language you mean biologically important
> functions?*

The term 'life history function' is a commonly used, direct term.

Here is a definition from Oxford:

"The life history of an organism is its pattern of survival and reproduction,
along with the traits that directly affect survival and the timing or amount of
reproduction. Rates of survival and reproduction can be estimated across age
classes, or across different stages in organisms with complex life cycles. Life
history traits include growth rate; age and size at sexual maturity; the
temporal pattern or schedule of reproduction; the number, size, and sex ratio of
offspring; the distribution of intrinsic or extrinsic mortality rates (e.g.,
patterns of senescence); and patterns of dormancy and dispersal."

To ensure the reader is clearly aware of which life-history behaviors/functions
we are referring to we provide specific examples within the sentence.

@@

We have moved this reference to later in the paragraph so the first
references are about Arctic work. While the example is from the
Antarctic, it is an important, early example of analysis using
bio-logger data to estimate haul-out behavior.

> *Do you mean age-class? The term life-history implies knowledge of
> previous behavioral efforts or occupancy of regions of interest.  
> Here you are specifically interested in age-class.*

We have edited to use the term 'demographic'

> Please help your reader understand why are you doing your analysis and not 
> offer a wide and jargon rich inventory of concepts that may be considered.

> Please keep your introduction focused on establishing the knowledge gap that 
> you are seeking to fill.   Please do not lead your reader to the numerous other 
> knowledge gaps that are not being addressed by your current study.

We have removed the sentence in question

> Please use the term "hauled-out".  Neither the biologger nor the aerial 
> survey is able to distinguish among the various seal behaviors that result in 
> them being hauled out and available for view by an aerial survey.

noted, and the terms 'hauled out', 'to haul out', etc is used consistently
throughout the manuscript

> You have indicate that you would consider sea ice extent.  Here you introduce 
> the concept of icescape.  Please introduce this concept earlier and indicate 
> how you will incorporate this into your analysis.

changed `icescape` to `sea-ice extent`

> Please use an active voice for clarity.

This sentence has been changed to:

"For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of 
collaborative investigations from 2005 through 2020."

> You summarize the years during which land-based studies applied bio-loggers.  
> Please do the same for the offshore deployments.

This sentence now reads:

"Ship-based capture events occurred during spring near the southern ice edge in
the Bering Sea between 2005 and 2018."

> Please provide the geographic context of your tag deployments and data 
> collection using a plot similar to that used by Jay et al. 
> (2012:  10.3354/meps10057 figure 2).  Please consider using a tabulation that 
> enables readers to understand the geographic context of the deployments 
> similar to that used by Jay et al. 2012, table 1.

We feel the spatial distribution of the actual transmitted data -- compared to
the deployment location -- is most informative and relevant for the audience's
understanding of how our data are distributed spatially and how well data
represents the range of each species. We provide this visually in Figure 3 and
adding an additional layer to that figure would to show deployments would be
an additional complication. To ensure the audience as a clear understanding
of our study area, we have added an additional figure with a map of the Bering
and Chukchi sea region that indicates generally where deployments started. We
have also added a more detailed figure to the supplemental section with each
location specifically marked.

> Please use active voice for all actions that you took in your data collection, 
> preparation and analysis.

Noted and addressed

> Please move this to the discussion.  This is a summary of your data findings, 
> not a description of your methods.

We can see the reviewer's point. However, we do not feel this topic is a summary 
of our data findings in this paper. Instead, our view
is that this information provides the context and rationale for our
methodological choice regarding bio-logger type and attachment methods. It is
also important regarding the analytic methods -- preference for flipper tag
haul-out data over those from a tag adhered to the dorsal/head. If we were to 
move these important points to the discussion, we worry those points would be 
left disconnected.

> This is a discussion of your methods.  This is not your methods.  Please state 
> how the data was collected in your Methods section and discuss how data 
> collection have been uncertain or biased in your discussion section.

> Please move to your discussion.

We appreciate this suggestion from the reviewer and we have moved large
portions of this paragraph to the discussion. However, some information
regarding age/sex class determination is left here so the reader has context
for Table 1.

> Please indicate how this was determined during data collection.

We have edited this sentence some to clearly communicate that field 
determination of reproductive status (and, thus ideal delineation of
sub-adult and adult age class) was not possible.

> Please allow your readers the opportunity to understand the geographic 
> context of these deployments through use of symbols that maybe used to 
> cross-reference the deployment locations in a map that indicates deployment 
> locations along with filtered movement tracks.

As previously addressed, there are now two figures available that provide
geographic context of the deployments. Figure 3 shows the geographic
distribution of the data from the deployments. The paper is not presenting or
discussing movement ecology and, therefore, we do not present filtered
movement tracks. We keep the focus on haul-out behavior data.

> Please consider bar plot similar to that used by Jay et al. 
> (2012: 10.3354/meps10057 figure 3), replacing the region with the seal species.

This figure has been updated with this suggestion in mind

> Please state clearly how you determined the end of a deployment.  Here you 
> indicate various elements of the haulout chronology, movement and dive 
> chronology that may have been used, however not reader could replicate your 
> methods base don this text.

We've added the following text to the paragraph:

"For example, a data record that ends with several consecutive
days (~10+ days) of 100% dry observations and with locations indicating the tag
was on land would be truncated to not include the final stretch of 100% dry
observations. The vast majority of deployments end with the device detaching in the 
water and the deployment end date is obvious. There is no perfect algorithm for
identifying deployment end dates and each deployment in question must be considered 
separately. While not perfect, we are confident our reliance on expert opinion and
examination of multiple data streams provides the best option."

> Deployment activity is known to disrupt normal pinniped behavior.  Did you 
> censure some period of data collection during and after the deployment to 
> ensure that that disruptions from the deployment activity did not bias your 
> analysis?

For these phocid species, we have no empirical data (or even anecdotal 
observational data) on how captures may or may not disrupt 'normal' behavior.
This is likely to be influenced by the capture method, how long the animal
was restrained, whether the animal was with a dependent pup, whether the
animal was molting, environmental variables such as wind and temperature, etc
etc. Given all these unknowns, we have not applied a censure to the beginning of
our deployment data. We can't 100% ensure this choice doesn't result in some
level of bias. However, given the large sample size and duration of deployments
we expect any bias imposed is negligible.

> This is a result.
> Please indicate the method that you used to distinguish "plausible" from 
> "implausible" telemetered seal behavior.  In your Results section, you may 
> indicate how method was able to identify occasions of implausible seal behavior 
> and any patterns that arose relative to tag model or attachment.

This sentence has been moved to the results section. Addtionally, we've provided
a brief explanation of implausible. For these devices, the percent-dry timeline
reported 100% dry for several weeks followed by an additional period of time
that was more typical of expected bearded seal behavior. During the extended dry
time the locations indicated movement of the tag that would not be simply
drifting ice. We also observed increasing transmission rates. The working (i.e.
unconfirmed guess) is that the wet/dry sensor surface became obscured by some
corrosive element such that when in the water conductivity was prevented and,
thus, recorded as dry.

> Please move to the Results section.
> Please move the discussion of how molt timing affects termination of tag 
> deployments to the discussion section.

Figure 1 (original) has been moved to the results section along with additional
descriptive text. We have also added mention to the Discussion section.

### 2023-02-04 analysis/paper/London_HauloutBeahvior.Rmd
additional edits re: rev #1

# knitr::opts_chunk$set(
#   tab.cap.style = "Table Caption",
#   tab.cap.pre = "Table ",
#   tab.cap.sep = ": "
#   )
# 
# knitr::opts_chunk$set(
#   fig.cap.style = "Image Caption",
#   fig.cap.pre = "Figure ",
#   fig.cap.sep = ": "
#   )

@@

Field identification of age class can be inexact, particularly when differentiating
subadults from adults. Sex as well as age class (non-dependent _young-of-the-year_,
sexually immature _subadults_, and mature _adults_) were estimated at the time
of deployment by various combinations of length, claw growth ridges
[@mclaren1958a], and pelage characteristics for some species. In the case of
ribbon seals, subadults often have less distinct ribbons than adults. Bearded
seal subadults will often have a spotted pattern in the pelage that is not seen
in adults. Spotted seal pelage cannot be used to reliably discern adults from
subadults. Despite these challenges, we feel the age classifications used in
this analysis are useful in testing for age-related effects on haul-out
behavior. Seals determined to be less than one year were classified as
young-of-the-year. Subadults are those seals likely greater than one year of age
but not yet sexually mature. Adults are individuals that are likely sexually
mature and older than four years. For those bio-loggers deployed on
young-of-the-year and transmitting into the next year, the age class was
advanced to subadult on 1 March of the following year. Table
\@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

Tags that fall off due to molt, attachment failure, or seal mortality and remain
on ice or land may continue to send data to satellites; i.e., a detached bio-logger 
that is dry (either on ice or land) will record and transmit data suggesting
the seal is hauled out. Therefore, end times of each deployment were identified by
examining bio-logger locations, percent-dry records, and dive behavior (if
available) to determine when bio-loggers ceased providing data consistent with
seal behavior. Data outside of the deployment start and end times were discarded
prior to analysis. After approximately 9 months, 7 devices deployed on the rear
flipper of bearded seals reported implausible hourly percent-dry data. All data
after the first instance of unrealistic values were censored from this analysis.
Figure \@ref(fig:dataCal) shows the distribution of all haul-out data across the
study season for each species. Observations for ribbon and spotted seals were
concentrated in the months of May and June due to the timing of deployment
(April and May) and the timing of molt (May and June). During molt, seals (and
their attached bio-loggers) spend more time out of the water and more data are
transmitted. Molt timing also impacts when many deployments end as any
bio-loggers adhered to the hair will fall off. Relative to the other species in
the study, there were fewer deployments of bio-loggers on bearded seals. This
resulted in fewer data observations overall and noticeably lower in numbers May
and June. The majority of bearded seal deployments started later in the summer and,
by May, bio-loggers had either fallen off or their batteries were depleted.

ggplot(data = seal_hours_tbl, aes(x = yday, y = n, fill = n))+
  geom_col() +
        scale_fill_gradientn(colors = rev(met.brewer("Hokusai2"))) +
        # scale_fill_viridis_c(option="mako") +
        scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
                           labels = c("01 March", "01 April", "01 May", "01 June", "01 July"),
                           expand = expansion(mult = c(0, 0))) +
        scale_y_continuous(expand = expansion(mult = c(0, 0))) +
  guides(fill = guide_colorbar(title.position = "bottom", 
                               title.hjust = 0.5, barwidth = 15,
                               barheight = 0.5)) +
  facet_grid(species ~ .) + 
  theme(axis.text.x = element_text(hjust = 0),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(fill = "seal-hours of bio-logger data") +
  ggtitle("Density of haul-out behavior observations varies from March through July",
          subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")

@@

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \@ref(fig:beardedHOCal)). Overall, bearded seals were
less likely to haul out and had a bi-modal distribution of haul-out probability
across the day. In addition to a peak around local solar noon, the bearded seal
model predicted additional haul-out activity around local midnight. In concert
with the lower magnitude of haul-out probability, bearded seal haul-out behavior
was also more protracted throughout the spring season compared to ribbon and
spotted seals.

@@

Unlike previous analyses of seal haul-out behavior in spotted and ribbon seals
(e.g. [@verhoef2010a], [@conn2014a]), we also investigated the influence of
sex-age class on haul-out probabilities for all species (not for bearded seals
because of low sample size). Although ribbon and spotted seals exhibited a
unimodal diel haul-out pattern generally centered around local solar noon, there
were key differences across species, age, and sex that match our understanding
from natural history descriptions of their ecological behavior. Spotted seals
are known to form triads during the breeding season where a female and dependent
pup are accompanied on the ice by a suitor male [@frost2018]. The male waits for
the female to wean the pup and enter estrus, and fends off any other potential
suitor males. Triad formation results in both males and females spending a large
portion of the day hauled out on ice and a protracted spring haul-out season for
both sexes. While females are still nursing the pup and not yet in estrus they
may be less inclined to interrupt their haul out and enter the water where the
suitor male would attempt mating. We see this reflected in the predicted
haul-out patterns, with both males and females exhibiting a broad distribution
of time out of the water throughout the solar day and the season. Ribbon seals
are not known to form triads and our model predicts a progression of increased
haul-out behavior with females starting earlier in the season than males.
Notably, female ribbon seals spend a large portion of the day in the water
during the pupping period, aligning with the hypothesis that ribbon seal females
continue foraging while nursing. Subadult ribbon and spotted seals begin
elevated haul-out behavior earlier in the spring and follow a pattern seen in
other phocids where yearlings and subadults molt first followed by adult females
and males [@reder2003, @thompson1987, @kirkman2003].

>>>>

@@

Sex as well as age class (non-dependent _young-of-the-year_, sexually immature
_subadults_, and mature _adults_) were estimated at the time of deployment by
various combinations of length, claw growth ridges [@mclaren1958a], and pelage
characteristics for some species. Seals determined to be less than one year were
classified as young-of-the-year. For those bio-loggers deployed on
young-of-the-year and transmitting into the next year, the age class was
advanced to subadult on 1 March of the following year. Subadults are those seals
likely greater than one year of age but less than four years. Adults are
individuals that are likely older than four years. Table \@ref(tab:tblDeploy)
provides a summary of these deployments and data received from them.

@@

Tags that fall off due to molt, attachment failure, or seal mortality and remain
on ice or land may continue to send data to satellites; i.e., a detached
bio-logger that is dry (either on ice or land) will record and transmit data
suggesting the seal is hauled out. Therefore, end times of each deployment were
identified by examining bio-logger locations, percent-dry records, and dive
behavior (if available) to determine when bio-loggers ceased providing data
consistent with seal behavior. For example, a data record that ends with several
consecutive days (~10+ days) of 100% dry observations and with locations
indicating the tag was on land would be truncated to not include the final
stretch of 100% dry observations. The vast majority of deployments end with the
device detaching in the water and the deployment end date is obvious. There is
no perfect algorithm for identifying deployment end dates and each deployment in
question must be considered separately. While not perfect, we are confident our
reliance on expert opinion and examination of multiple data streams provides the
best option. Data outside of the deployment start and end times were discarded
prior to analysis.

@@

Figure \@ref(fig:dataCal) shows the distribution of all
haul-out data across the study season for each species. Observations for ribbon
and spotted seals were concentrated in the months of May and June due to the
timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the
water and more data are transmitted. Molt timing also impacts when many
deployments end as any bio-loggers adhered to the hair will fall off. Relative
to the other species in the study, there were fewer deployments of bio-loggers
on bearded seals. This resulted in fewer data observations overall and
noticeably lower in numbers May and June. The majority of bearded seal
deployments started later in the summer and, by May, bio-loggers had either
fallen off or their batteries were depleted.

ggplot(data = seal_hours_tbl, aes(x = yday, y = n, fill = n))+
  geom_col() +
        scale_fill_gradientn(colors = met.brewer("VanGogh3")) +
        # scale_fill_viridis_c(option="mako") +
        scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
                           labels = c("01 March", "01 April", "01 May", "01 June", "01 July"),
                           expand = expansion(mult = c(0, 0))) +
        scale_y_continuous(expand = expansion(mult = c(0, 0))) +
  guides(fill = guide_colorbar(title.position = "bottom", 
                               title.hjust = 0.5, barwidth = 15,
                               barheight = 0.5)) +
  facet_grid(species ~ .) + 
  theme(axis.text.x = element_text(hjust = 0),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(fill = "seal-hours of bio-logger data") +
  ggtitle("Density of haul-out behavior observations varies from March through July",
          subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \@ref(fig:beardedHOCal)). Additionally, after
approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
reported implausible hourly percent-dry data (100% dry for several weeks but
indicative of movement and increasing transmission rates (see @boveng2013)). All
data after the first instance of unrealistic values were censored from this
analysis. Overall, bearded seals were less likely to haul out and had a bi-modal
distribution of haul-out probability across the day. In addition to a peak
around local solar noon, the bearded seal model predicted additional haul-out
activity around local midnight. In concert with the lower magnitude of haul-out
probability, bearded seal haul-out behavior was also more protracted throughout
the spring season compared to ribbon and spotted seals.

@@

Unlike previous analyses of seal haul-out behavior in spotted and ribbon seals
(e.g. [@verhoef2010a], [@conn2014a]), we also investigated the influence of
sex-age class on haul-out probabilities for both species (not for bearded seals
because of low sample size). Field identification of age class can be inexact,
particularly when differentiating subadults from adults. In the case of ribbon
seals, subadults often have less distinct ribbons than adults. Spotted seal
pelage cannot be used to reliably discern adults from subadults. Despite these
challenges, we feel the age classifications used in this analysis are useful in
testing for age-related effects on haul-out behavior.

Although ribbon and spotted seals exhibited a
unimodal diel haul-out pattern generally centered around local solar noon, there
were key differences across species, age, and sex that match our understanding
from natural history descriptions of their ecological behavior. Spotted seals
are known to form triads during the breeding season where a female and dependent
pup are accompanied on the ice by a suitor male [@frost2018]. The male waits for
the female to wean the pup and enter estrus, and fends off any other potential
suitor males. Triad formation results in both males and females spending a large
portion of the day hauled out on ice and a protracted spring haul-out season for
both sexes. While females are still nursing the pup and not yet in estrus they
may be less inclined to interrupt their haul out and enter the water where the
suitor male would attempt mating. We see this reflected in the predicted
haul-out patterns, with both males and females exhibiting a broad distribution
of time out of the water throughout the solar day and the season. Ribbon seals
are not known to form triads and our model predicts a progression of increased
haul-out behavior with females starting earlier in the season than males.
Notably, female ribbon seals spend a large portion of the day in the water
during the pupping period, aligning with the hypothesis that ribbon seal females
continue foraging while nursing. Subadult ribbon and spotted seals begin
elevated haul-out behavior earlier in the spring and follow a pattern seen in
other phocids where yearlings and subadults molt first followed by adult females
and males [@reder2003, @thompson1987, @kirkman2003].
